---
title: "Weather events and their impacts in the USA"
author: "Mathias Barat"
date: "21/09/2020"
output: html_document
---

# 1 - Synopsis

The present study is adressing several questions linked to the impacts dealt by weather events in the contemporary USA. We will focus mainly on the following points:

  - Harmfulness of the events toward the population health
  - Economic impacts of the events
  
The study is based on the NOAA Storm Database (URLs given hearafter) :

```{r}
url_storm_data <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
url_storm_doc <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
url_storm_faq <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"
```

The present study will answer to the following points:

- Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

- Across the United States, which types of events have the greatest economic consequences?

# 2 - DATA-PROCESSING

### PREPARATION

```{r}
library(ggplot2)
library(plyr)
library(dplyr)
```


### DATA LOADING

```{r}

if (!dir.exists("data/")){dir.create("data/")}
if (!file.exists("data/storm_data.csv")){
  download.file(url_storm_data, destfile = "data/storm_data.csv")
  download.file(url_storm_doc, destfile = "data/storm_doc.pdf")
  download.file(url_storm_faq, destfile = "data/storm_faq.pdf")
}

storm_data <- read.csv("data/storm_data.csv")

```

### DATA STRUCTURE

Here is some information about the RAW dataset :

```{r}
dim(storm_data)
```
There is 902297 observations for 37 variables.

```{r}
head(storm_data)
```
```{r}
str(storm_data)
```

### Extract Data Variables of Interests

We can see that a lot of variables are not useful for our study (empty or not relevant). I am capitalizing hereafter the variables I will keep for my analysis:

### Health Variables:
  - FATALITIES
  - INJURIES

### Economics Variables:
  - PROPDMG
  - PROPDMGEXP
  - CROPDMG
  - CROPDMGEXP
  
### Study Variables:
  - STATE
  - EVTYPE


```{r}
Cols_to_keep <- c("STATE", "EVTYPE","FATALITIES", "INJURIES", "PROPDMG","PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
storm_data_wip <- storm_data[ ,Cols_to_keep]
```

### Check NA

```{r}
sum(is.na(storm_data_wip))
```
There is no NA in the filtered dataset

### Clean EVTYPE

In order to be more accurate, we will regroup the EVTYPE has too much similar observations which can be packed

Creation of storm_data_wip$EVENT
```{r}
storm_data_wip$EVENT <- "OTHER"
storm_data_wip$EVENT[grep("HAIL", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "HAIL"
storm_data_wip$EVENT[grep("WIND", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "WIND"
storm_data_wip$EVENT[grep("TORNADO", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
storm_data_wip$EVENT[grep("FLOOD", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
storm_data_wip$EVENT[grep("LIGHTNING", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "LIGHTNING"
storm_data_wip$EVENT[grep("SNOW", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "SNOW"
storm_data_wip$EVENT[grep("RAIN", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "RAIN"
storm_data_wip$EVENT[grep("STORM", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "STORM"
storm_data_wip$EVENT[grep("TSTM", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "THUNDERSTORM"
storm_data_wip$EVENT[grep("THUNDERSTORM", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "THUNDERSTORM"
storm_data_wip$EVENT[grep("FIRE", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "FIRE"
storm_data_wip$EVENT[grep("FOG", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "FOG"
storm_data_wip$EVENT[grep("BLIZZARD", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "BLIZZARD"
storm_data_wip$EVENT[grep("HEAT", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "HEAT"
storm_data_wip$EVENT[grep("FUNNEL CLOUD", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "FUNNEL CLOUD"
storm_data_wip$EVENT[grep("COLD", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "COLD"
storm_data_wip$EVENT[grep("WATERSPOUT", storm_data_wip$EVTYPE, ignore.case = TRUE)] <- "WATERSPOUT"
```

### Dollar conversion

As the amount of damages are expressed with expression such as K, M, B in the ...EXP columns.
We need to create first columns with actual dollars:

```{r}
storm_data_wip$PROPDMGEXP <- as.character(storm_data_wip$PROPDMGEXP)
storm_data_wip$PROPDMGEXP[is.na(storm_data_wip$PROPDMGEXP)] <- 0
storm_data_wip$PROPDMGEXP[!grepl("K|M|B", storm_data_wip$PROPDMGEXP, ignore.case = TRUE)] <- 0
storm_data_wip$PROPDMGEXP[grep("K", storm_data_wip$PROPDMGEXP, ignore.case = TRUE)] <- "3"
storm_data_wip$PROPDMGEXP[grep("M", storm_data_wip$PROPDMGEXP, ignore.case = TRUE)] <- "6"
storm_data_wip$PROPDMGEXP[grep("B", storm_data_wip$PROPDMGEXP, ignore.case = TRUE)] <- "9"
storm_data_wip$PROPDMGEXP <- as.numeric(as.character(storm_data_wip$PROPDMGEXP))
storm_data_wip$PROPDMG <- as.numeric(as.character(storm_data_wip$PROPDMG))
storm_data_wip$PROPDMG.DOLLAR <- storm_data_wip$PROPDMG * 10^storm_data_wip$PROPDMGEXP

storm_data_wip$CROPDMGEXP <- as.character(storm_data_wip$CROPDMGEXP)
storm_data_wip$CROPDMGEXP[is.na(storm_data_wip$CROPDMGEXP)] <- 0 # NA's considered as dollars
storm_data_wip$CROPDMGEXP[!grepl("K|M|B", storm_data_wip$CROPDMGEXP, ignore.case = TRUE)] <- 0
storm_data_wip$CROPDMGEXP[grep("K", storm_data_wip$CROPDMGEXP, ignore.case = TRUE)] <- "3"
storm_data_wip$CROPDMGEXP[grep("M", storm_data_wip$CROPDMGEXP, ignore.case = TRUE)] <- "6"
storm_data_wip$CROPDMGEXP[grep("B", storm_data_wip$CROPDMGEXP, ignore.case = TRUE)] <- "9"
storm_data_wip$CROPDMGEXP <- as.numeric(as.character(storm_data_wip$CROPDMGEXP))
storm_data_wip$CROPDMG.DOLLAR <- storm_data_wip$CROPDMG * 10^storm_data_wip$CROPDMGEXP
```

### Health Data Preparation

```{r}
Mycolnames <- c("EVENT", "TOTAL", "TYPE") 
most_dangerous_events.injuries <- aggregate( INJURIES ~ EVENT , storm_data_wip, sum)
most_dangerous_events.injuries$TYPE <- "INJURIES"
colnames(most_dangerous_events.injuries) <- Mycolnames

most_dangerous_events.fatalities <- aggregate( FATALITIES ~ EVENT , storm_data_wip, sum)
most_dangerous_events.fatalities$TYPE <- "FATALITIES"
colnames(most_dangerous_events.fatalities)  <- Mycolnames

most_dangerous_events <- rbind(most_dangerous_events.fatalities, most_dangerous_events.injuries)

```

### Damage Data Preparation

```{r}

damages.prop <- aggregate( PROPDMG.DOLLAR ~ EVENT , storm_data_wip, sum)
damages.prop$TYPE <- "PROP"
colnames(damages.prop) <- Mycolnames

damages.crop <- aggregate( CROPDMG.DOLLAR ~ EVENT , storm_data_wip, sum)
damages.crop$TYPE <- "CROP"
colnames(damages.crop)  <- Mycolnames

damages <- rbind(damages.prop, damages.crop)

```

# 3 - RESULTS

### EVENTS ANALYSIS

```{r}
frequent_events <- as.data.frame(sort(table(storm_data_wip$EVENT), decreasing=TRUE)[1:30])
```
#### Global numbers for health:

Total amount of fatalities in the dataset:
```{r}
sum(storm_data_wip$FATALITIES)
```
Total amount of injuries in the dataset:
```{r}
sum(storm_data_wip$INJURIES)
```
## GEOGRAPHICAL ANALYSIS

```{r}
most_impacted_states.events <- as.data.frame(with(storm_data_wip, table(EVENT,STATE)))
most_impacted_states.events <- most_impacted_states.events[order(most_impacted_states.events$Freq, decreasing = TRUE),]
```
```{r}
ggplot(most_impacted_states.events[1:100,] , aes(reorder(STATE, -Freq) ,Freq, fill = EVENT)) +
  geom_bar(stat='identity') +
  ylab("Count of Events")+
  xlab("States")+
  theme(axis.text.x=element_text(angle=90,hjust = 1))+
  ggtitle("Top 100 Most Frequent Events by States", subtitle = "Based on the NOAA Storm Database - 1950 to nowadays")
```

## HARMFULNESS ANALYSIS


```{r}

ggplot(most_dangerous_events , aes(EVENT, TOTAL , fill = TYPE)) +
  geom_bar(stat='identity') +
  ylab("Injuries + Fatalities")+
  xlab("Weather Events")+
  theme(axis.text.x=element_text(angle=90,hjust = 1))+
  ggtitle("People Deaths and Injuries caused by Weather Events in USA", subtitle = "Based on the NOAA Storm Database - 1950 to nowadays")

```

### Sum-up

Tornados are far more the most dangerous weather events USA is dealing with. They are responsible of more than 90000 injuries and more than 5000 fatalities.

Heat has a high amount of fatalities as it has killed more than 3000 people since 1950.
Floods are closing the top killers with around 1500 deaths.

## DAMAGE ANALYSIS

```{r}

ggplot(damages , aes(EVENT, TOTAL , fill = TYPE)) +
  geom_bar(stat='identity') +
  ylab("Total Damage in Dollars")+
  xlab("Weather Events")+
  theme(axis.text.x=element_text(angle=90,hjust = 1))+
  ggtitle("People Deaths and Injuries caused by Weather Events in USA", subtitle = "Based on the NOAA Storm Database - 1950 to nowadays")

```
### Sum-up

Floods caused more than 170 Billion dollars of damages since 1950.